---
title: "Pokemon"
author: "John Cruz"
date: "2023-03-01"
output: pdf_document
url_color: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

National Pokedex

[Pokemon Stats](https://www.serebii.net/pokemon/nationalpokedex.shtml)

[Pokemon Types](https://m.bulbapedia.bulbagarden.net/wiki/List_of_Pok%C3%A9mon_by_National_Pok%C3%A9dex_number)

A few kinds of analysis that could be done would be:

to look at which types of Pokémon tend to have the highest base stats
to find the 6 Pokémon with the highest base speed
to observe whether Pokémon with a single type have higher base stats than Pokémon with two types
to observe the frequency of Pokémon by type
to find the 10 rarest Pokémon abilities, i.e. the abilities that the fewest number of Pokémon have access to


```{r library, warning=FALSE, message=FALSE}
library(tidyverse)
library(rvest)
library(xml2)
library(janitor)
```

```{r read-url-pokemon-stats}
url <- "https://www.serebii.net/pokemon/nationalpokedex.shtml"

web_table <- read_html(url) 

# use XML to account for <br> with abilities and add '\n'
xml_find_all(web_table, ".//br") |> 
  xml_add_sibling("p", "\n")
xml_find_all(web_table, ".//br") |> 
  xml_remove() 

web_table <- 
  web_table |> 
  html_element('.dextable') |> 
  html_table()

pokemon_stats <- as.data.frame(web_table)

```



```{r tidy-pokemon-stats}

# drop null values if Pokemon name is N/A
stats_df <- 
  pokemon_stats |> 
  drop_na(4)

# drop first row (duplicate header) and second column (pic)
stats_df <- stats_df[-1,-2]

# set column headers from first row and clean names
stats_df <- 
  stats_df |> 
  row_to_names(row_number = 1) |> 
  clean_names()

# shift pokemon names, etc to left by 1 column
stats_df[c(2:10)] = stats_df[, c(3:11)]

# drop 'na' column
stats_df <- 
  stats_df |> 
  select(!c(na, type))

# split multiple abilities into long format based on created '\n'
stats_df <- 
  stats_df |> 
  separate_longer_delim(abilities, delim = "\n")

# change to pokemon number
stats_df$no <-
  parse_number(stats_df$no)

```



```{r read-url-pokemon-types}
url <- "https://bulbapedia.bulbagarden.net/wiki/List_of_Pok%C3%A9mon_by_National_Pok%C3%A9dex_number"

web_table <- read_html(url) 

# use XML to account for <br> and replace with '\n'
xml_find_all(web_table, ".//br") |> 
  xml_add_sibling("p", "\n")
xml_find_all(web_table, ".//br") |> 
  xml_remove() 

web_table <- 
  web_table |> 
  html_element('body') |> 
  html_table()

pokemon_types <- as.data.frame(web_table)
```



```{r tidy-pokemon-types, warning=FALSE}

# drop null values if Pokemon name is N/A
types_df <- 
  pokemon_types |> 
  drop_na(2)

# drop unnecessary columns
types_df <- 
  types_df[, 1:5]

# set column headers from first row and clean names
types_df <- 
  types_df |> 
  row_to_names(row_number = 1) |> 
  clean_names()

# change to pokemon number
types_df$ndex <-
  parse_number(types_df$ndex)

# drop N/A or zero (0) while keeping only distinct pokemon numbers
types_df <- 
  types_df |> 
  drop_na() |> 
  filter(ndex != 0) |> 
  distinct(ndex, .keep_all=TRUE)

# within same pokemon number, replace repeated types with N/A
types_df <- 
  types_df |> 
  mutate(type_2 = if_else(type_2 != type, type_2, NA)) |> 
  select(-c(2)) |> 
  rename(no = ndex)

# melt both type columns into one column
temp1 <- 
  types_df |> 
  select(1:3)

temp2 <- 
  types_df |> 
  select(1,2,4) |> 
  rename(type = type_2)

types_df <- 
  temp1 |> 
  full_join(temp2) |>
  drop_na() |> 
  select(!pokemon) |> 
  arrange(no)

```



```{r merged-stats-types}
stats_types_df <-  
  stats_df |> 
  inner_join(types_df) |> 
  relocate(type, .after = name)
```



```{r plot}
stats_types_df |> 
  group_by(type) |> 
  summarise(count = n()) |> 
  arrange(desc(count)) |> 
  ggplot(aes(x = count, y = type)) +
  geom_bar(stat = 'identity')
  
```
