---
title: "Pokemon"
author: "John Cruz"
date: "2023-03-01"
output: 
  html_document: default
  pdf_document: default
url_color: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

The Serebii website provides a list, called the National Pokedex, of all the Pokemon in all the games. The table breaks down each Pokemon with a unique ID number, name, type, abilities and base stats such as attack and defense.  

**Data Sources:** [Pokemon Stats](https://www.serebii.net/pokemon/nationalpokedex.shtml)

Lets determine a frequency chart of which Pokemon fall into which types they are. 

## Required Libraries

```{r library, warning=FALSE, message=FALSE}
library(tidyverse)
library(rvest)
library(xml2)
library(janitor)
```

---

## Web Scrape Table into Data Frame

At first, when the table was web scraped, it brought in the values into a Data Frame, however, I quickly realized that the abilities listed multiple ones and the *rvest()* package cleaned it up by adding a space between them. This causes issues as some abilities are multiple words so to string manipulate them, I needed to account for the built-in HTML line breaks (<br>). To do this, I used XML to add an intentional new line character (\n), so that I could be able to break them apart later. 

[Using XML](https://stackoverflow.com/questions/30921626/can-rvest-keep-inline-html-tags-such-as-br-using-html-table)

```{r read-url-pokemon-stats}
url <- "https://www.serebii.net/pokemon/nationalpokedex.shtml"

web_table <- read_html(url) 

# use XML to account for <br> with abilities and add '\n'
xml_find_all(web_table, ".//br") |> 
  xml_add_sibling("p", "\n")
xml_find_all(web_table, ".//br") |> 
  xml_remove() 

web_table <- 
  web_table |> 
  html_element('.dextable') |> 
  html_table()

pokemon_stats <- as.data.frame(web_table)

knitr::kable(head(pokemon_stats))

```

---

## Tidy Up Data

Every other row that was brought into the Data Frame had *N/A* values. I removed these rows where if any Pokemon didn't have a name, it was dropped. The columns between *Name* and the *Base Stats* also were shifted to the right by one column. Using a vector, I moved the columns to the appropriate alignment and followed it by using the *janiotr* package to clean up column names. To account for multiple abilities, the data frame was then unpivoted into a longer format for each Pokemon. 

**Note:** Pokemon types are missing and the column was dropped. It will be explained in the following section.

```{r tidy-pokemon-stats}

# drop null values if Pokemon name is N/A
stats_df <- 
  pokemon_stats |> 
  drop_na(4)

# drop first row (duplicate header) and second column (pic)
stats_df <- stats_df[-1,-2]

# set column headers from first row and clean names
stats_df <- 
  stats_df |> 
  row_to_names(row_number = 1) |> 
  clean_names()

# shift pokemon names, etc to left by 1 column
stats_df[c(2:10)] = stats_df[, c(3:11)]

# drop 'na' column
stats_df <- 
  stats_df |> 
  select(!c(na, type))

# split multiple abilities into long format based on created '\n'
stats_df <- 
  stats_df |> 
  separate_longer_delim(abilities, delim = "\n")

# change to pokemon number
stats_df$no <-
  parse_number(stats_df$no)

knitr::kable(head(stats_df))

```

---

## Pokemon Types Missing

While working on the Data Frame, I realized the Pokemon types were missing. Looking into the HTML code, they types were not text, but images with no conventional tag name for each one. Beause of this, I used another website that provides that information and web scraped similarly into another Data Frame.

```{r read-url-pokemon-types}
url <- "https://bulbapedia.bulbagarden.net/wiki/List_of_Pok%C3%A9mon_by_National_Pok%C3%A9dex_number"

web_table <- read_html(url) 

# use XML to account for <br> and replace with '\n'
xml_find_all(web_table, ".//br") |> 
  xml_add_sibling("p", "\n")
xml_find_all(web_table, ".//br") |> 
  xml_remove() 

web_table <- 
  web_table |> 
  html_element('body') |> 
  html_table()

pokemon_types <- as.data.frame(web_table)

knitr::kable(head(pokemon_types))
```

Since each Pokemon has a unique identifier, I created the data frame to relate this ID to its types. 

```{r tidy-pokemon-types, warning=FALSE, message=FALSE}

# drop null values if Pokemon name is N/A
types_df <- 
  pokemon_types |> 
  drop_na(2)

# drop unnecessary columns
types_df <- 
  types_df[, 1:5]

# set column headers from first row and clean names
types_df <- 
  types_df |> 
  row_to_names(row_number = 1) |> 
  clean_names()

# change to pokemon number
types_df$ndex <-
  parse_number(types_df$ndex)

# drop N/A or zero (0) while keeping only distinct pokemon numbers
types_df <- 
  types_df |> 
  drop_na() |> 
  filter(ndex != 0) |> 
  distinct(ndex, .keep_all=TRUE)

# within same pokemon number, replace repeated types with N/A
types_df <- 
  types_df |> 
  mutate(type_2 = if_else(type_2 != type, type_2, NA)) |> 
  select(-c(2)) |> 
  rename(no = ndex)

# melt both type columns into one column
temp1 <- 
  types_df |> 
  select(1:3)

temp2 <- 
  types_df |> 
  select(1,2,4) |> 
  rename(type = type_2)

types_df <- 
  temp1 |> 
  full_join(temp2) |>
  drop_na() |> 
  select(!pokemon) |> 
  arrange(no)

knitr::kable(head(types_df))

```

## Merge Pokemon Stats and Types Tables

Using a similar way of SQL joins, I merged the Pokemon to its types and stats. 

```{r merged-stats-types, message=FALSE, warning=FALSE}
stats_types_df <-  
  stats_df |> 
  inner_join(types_df) |> 
  relocate(type, .after = name)

knitr::kable(head(stats_types_df))
```

---

## Pokemon Type Frequency

The frequency chart shows us that *Water, Normal and Grass* types are the most common among all the Pokemon. One caveat to point out, is that a Pokemon could have up to two different types. This means that a Pokemon can appear twice in different categories as a result in here. 

```{r plot}
grouped_type <-  
  stats_types_df |> 
  distinct(name, type) |> 
  group_by(type) |> 
  summarise(count = n()) |> 
    arrange(desc(count))

grouped_type |> 
  ggplot(aes(x = count, y = reorder(type, count))) +
  geom_bar(stat = 'identity')

knitr::kable(grouped_type)
```
